---
title: "GDP Per Capita, Price Levels and Population For European Countries"
output:
  pdf_document: default
  html_notebook: default
---
```{r}
knitr::opts_chunk$set(warning = FALSE, error = FALSE, message = FALSE,
                      fig.align = 'center', fig.height = 5, fig.width = 5, tidy=TRUE)
```


## Introduction 

For our project we were particularly interested in the relation between GDP per capita and price levels in different European countries. We got our data from the Eurostat website. In the code below we begin by importing our dataset from the Eurostat website and cleaning it by removing the columns we do not need. 

```{r, include=FALSE, echo=FALSE}
library(eurostat)
library(tidyverse)
library(corrplot)
library(dplyr)
library(sf)
library(ggplot2)
```


```{r, echo=FALSE}
#read data 
price_data <- read_csv("data/comparative-price-levels.csv",  show_col_types = FALSE)
gdp_data <- read_csv("data/gdp-per-capita.csv",  show_col_types = FALSE)
pop_data <- read_csv("data/population.csv",  show_col_types = FALSE)

##CLEAN PRICE DATA 
#DELETE EXTRA COLUMNS 
price_data <- subset(price_data, select = -DATAFLOW )
price_data <- subset(price_data, select = -OBS_FLAG )
price_data <- subset(price_data, select = -freq )
price_data <- subset(price_data, select = -na_item )
price_data <- subset(price_data, select = -ppp_cat )

#RENAME COLUMNS 
names(price_data)[names(price_data) == "OBS_VALUE"] <- "price_levels"

##CLEAN GDP DATA
#DELETE EXTRA COLUMNS
gdp_data <- subset(gdp_data, select = -DATAFLOW )
gdp_data <- subset(gdp_data, select = -OBS_FLAG )
gdp_data <- subset(gdp_data, select = -freq )
gdp_data <- subset(gdp_data, select = -na_item )
gdp_data <- subset(gdp_data, select = -ppp_cat )

#RENAME CERTAIN COLUMNS
names(gdp_data)[names(gdp_data) == "OBS_VALUE"] <- "GDP"


##CLEAN POPULATION DATA 
pop_data <- subset(pop_data, select = -freq )
pop_data <- subset(pop_data, select = -OBS_VALUE )
pop_data <- subset(pop_data, select = -OBS_FLAG )
pop_data <- subset(pop_data, select = -indic_de )
#pop_data <- subset(pop_data, select = -LAST UPDATE )
pop_data <- subset(pop_data, select = -DATAFLOW )

#RENAME CERTAIN COLUMNS
names(pop_data)[names(pop_data) == "OBS_VALUE"] <- "Population"


```


```{r}
##MERGE DATASETS
total <- merge(gdp_data,price_data,by=c("TIME_PERIOD","geo"))

total <- merge(total,pop_data,by=c("TIME_PERIOD","geo"))

total <- na.omit(total)

total <- subset(total, select = -c(`LAST UPDATE`))
total <- subset(total, select = -c(`LAST UPDATE.x`))
total <- subset(total, select = -c(`LAST UPDATE.y`))

total <- total %>% filter(geo !="EU27_2020")
total <- total %>% filter(geo !="EA19")
total <- total %>% filter(geo !="EA20")
total <- total %>% filter(geo !="US")


total <- inner_join(total, eu_countries, by = join_by(geo == code))

total <- subset(total, select = -label )
```


## Average Price Levels in Different European Countries Between 2011 and 2022 
Below we create an initial plot showing the average price levels in different European countries between the years 2012 to 2022. As seen below it is evident that Denmark (DK) has the highest average price levels while Bulgaria has the lowest average price levels.  

```{r}
#CREATE AN INITIAL PRICE PLOT
ggplot(total, aes(x = 
                    reorder(x=geo,
                            X=price_levels),
                            y=price_levels)) + xlab("Countries") + ylab("Price Levels") +
  geom_bar(stat = "summary", fun="mean") + coord_flip() + theme(legend.position = "none")
```

## Price and GDP Levels in Different European Regions

We decided to get an approximation of price levels in different European regions to see if a certain region of the continent performs better in these metrics than others. As there is no defined regions, we used ChatGPT to help us with dividing the different parts of Europe. This is shown below.

```{r}
##CREATE A REGION BASED COLUMN

western_europe <- c("BE", "DK", "DE", "IE", "FR", "IT", "LU", "NL", "PT", "ES")
northern_europe <- c("EE", "LV", "LT", "FI", "SE")                           
eastern_europe <- c("BG", "CZ", "HR", "HU", "PL", "RO", "SK","SI")            
southern_europe <- c("EL", "CY", "MT", "AT") 

total$regions <- ifelse(total$geo %in% western_europe,
                  "Western Europe",
                  ifelse(total$geo %in% eastern_europe, "Eastern Europe",
                  ifelse(total$geo %in% southern_europe, "Southern Europe","Northern Europe")))

```

The plot below shows the average price levels in different European regions over the years 2011 to 2022. As we can see, Denmark performs the best on average and Bulgaria performs the worst. 

```{r}

##CREATE A REGION BASED PRICE PLOT 

plot_1 <- ggplot(total, aes(x = 
                    reorder(x=geo,
                            X=price_levels),
                            y=price_levels, 
                    order=regions,
                            fill=regions, color = regions, shape=regions)) +  labs(y = "Price levels", x = "Countries")+ geom_point(stat = "summary", fun="mean", size=3)+ coord_flip()+  theme_minimal() 


plot_1
```
```{r}


arrange(total['price_levels'])

```

The plot below shows the average GDP per capita in different European regions over the years 2011 to 2022. As we can see, Luxembourg performs the best on average and Bulgaria performs the worst. We can understand this by the fact that GDP is calculated per capita and Luxembourg has one of the smallest populations.  

```{r}

#CREATE A REGION BASED GDP PLOT 
plot_2 <- ggplot(total, aes(x = 
                    reorder(x=geo,
                            X=price_levels), # this X is what you want to order by 
                            y=GDP,
                            order=regions,
                            fill = regions, color = regions, shape=regions)) +  labs(y = "Gross Domestic Product Per Capita", x = "Countries", fill = "REGION") + geom_point(stat = "summary", fun="mean", size=3) + coord_flip()  + theme_minimal() +
  theme(legend.position="none")


plot_2





```

# Custom ordered bar plot in ascending order based on values
custom_order_plot_ascending_values <- ggplot(data, aes(x = reorder(category, value, 
												FUN = function(x) sum(x)), 
																y = value)) +
geom_bar(stat = "identity", fill = "skyblue") +
labs(title = "Custom Ordered Bar Plot (Ascending by Values)")

# Display the custom ordered plot in ascending order based on values
print(custom_order_plot_ascending_values)




```{r}
library(gridExtra)

# Create the heading with smaller text size
heading <- textGrob("Scatter Plot for GDP per capita and Price Levels for European Countries separated by Region for 2022",
                    gp = gpar(fontsize = 10)) 


combined_plots  <- grid.arrange(plot_1, plot_2, ncol=2, top = heading)



# Arrange the GDP into same order as the Price levels 
# CHange the colors 


# Add caption
#caption <- "Scatter Plot for GDP per capita and Price Levels for European Countries separated by region"
#grid.text(caption, x = unit(0.5, "npc"), y = unit(0.05, "npc"), just = "center", gp = gpar(fontsize = 10))

# Print combined plot with caption
print(combined_plots)

```

```{r}
# Default order
total$regions
```




# add gdp per capita 






## Geospatial Plots 

Here we visualise the data onto an actual map. As shown below, Luxembourg and the UK have the highest GDP.

```{r}


#SHP_27
SHP_0 <- get_eurostat_geospatial(resolution = 10, 
                                 nuts_level = 0, 
                                 year = 2016)

EU28 <- eu_countries %>% 
  select(geo = code, name)
SHP_27 <- SHP_0 %>% 
  select(geo = NUTS_ID, geometry) %>% 
  inner_join(EU28, by = "geo") %>% 
  arrange(geo) %>% 
  st_as_sf()

gdp_map <- total %>% 
  filter(TIME_PERIOD == 2022) %>% 
  select(geo, GDP) %>% 
  inner_join(SHP_27, by = "geo") %>% 
  st_as_sf()

gdp_map %>% 
  ggplot(aes(fill = GDP)) +
  geom_sf() +
  scale_fill_continuous(type = "viridis", na.value ="black") +
  scale_x_continuous(limits = c(-10, 35)) +
  scale_y_continuous(limits = c(35, 65)) +
  theme_bw() + 
  labs(caption="Varying GDP per European country in 2022")


```
We can also see that Denmark, Luxembourg and the UK have the highest price levels on average. 

```{r}

price_map <- total %>% 
  filter(TIME_PERIOD == 2022) %>% 
  select(geo, price_levels) %>% 
  inner_join(SHP_27, by = "geo") %>% 
  st_as_sf()

price_map %>% 
  ggplot(aes(fill = price_levels)) +
  geom_sf() +
  scale_fill_continuous(type = "viridis") +
  scale_x_continuous(limits = c(-10, 35)) +
  scale_y_continuous(limits = c(35, 65))  + theme_void() + 
  labs(caption="Varying price levels per European country")
```



```{r, include=FALSE, echo=FALSE}

#region_map <- total %>% 
 # filter(TIME_PERIOD == 2022) %>% 
 # select(regions, price_levels) %>% 
 # inner_join(SHP_27, by = "geo") %>% 
 # st_as_sf()


#region_map %>% 
 # ggplot(aes(fill = regions)) +
 # geom_sf() +
  #scale_fill_continuous(type = "viridis") +
 # scale_x_continuous(limits = c(-10, 35)) +
 # scale_y_continuous(limits = c(35, 65)) +
 # theme_void() + 
 # labs(caption="Varying price levels per European region")

```


## Spaghetti Plot 

The spaghetti plot below illustrates the GDP per capita over time for the European countries. It is clear that the GDP for Ireland increases over time dramatically and the GDP for Luxembourg is the highest which makes sense since it is a rich small country and this is GDP per capita. 


```{r}


total$new_name = ifelse (total$name == "Ireland", "Ireland", ifelse(total$name == "Luxembourg", "Luxembourg", "other"))
```


```{r}
# spagetti plot using total and gdp 
ggplot(total, aes(x = TIME_PERIOD, y = GDP, group = name, color = new_name)) + 
  geom_line() +
  labs(title = "Spaghetti Plot",
       x = "Year",
       y = "Value",
       caption = "Spaghetti plot showing the progression of GDP from 2012 to 2022") +
  theme_minimal() +
  scale_color_manual(name = "Country",
                     values = c("Ireland" = "green", "Luxembourg" = "lightblue", "Other" = "grey"))
```

## Scatterplot 

The figure below displays the scatter plot for the respective GDP value and price level for the European countries. The GDP value is on the x-axis and price level is on the y-axis. The colours represent the regions of Europe. Each dot on the scatter plot represents the price level and GDP for a specific country at a certain year. There is a positive trend between GDP and price level as one would expect. 


```{r}
#new_total<- total$TIME_PERIOD=="2022"
new_total <- subset(total, TIME_PERIOD == 2022)

ggplot(new_total, aes(x=GDP, y=price_levels, color=regions)) + geom_point() + labs(caption = "Scatter plot showing GDP vs. price levels for the year 2022 by region")
```


```{r,  include=FALSE, echo=FALSE}
# Scatter plot 

new_data <- data.frame(label = rep(NA, 297))

new_data$label <- total$label
new_data$regions <- total$regions

average_price_by_country <- aggregate(price_levels ~ geo, data = total, FUN = mean, na.rm = TRUE)

##merge price 

merged_df <- merge(total, new_data, by = "regions", all = TRUE)

new_data <- data.frame(label = unique(total$regions))

# Calculate average price levels for each country
average_price_by_country <- aggregate(price_levels ~ regions, data = total, FUN = mean, na.rm = TRUE)

print(new_data)
#print(average)
# Merge total and new_data based on the label column
merged_df <- merge(new_data, average_price_by_country, by.x = "label", by.y = "regions", all.x = TRUE)

# Rename the columns as needed
names(merged_df)[1] <- "regions"
names(merged_df)[2] <- "Average_Price"

```


```{r}
##merge gdp 

new_data <- data.frame(label = unique(total$regions))

# Calculate average price levels for each country
average_GDP_by_country <- aggregate(GDP ~ regions, data = total, FUN = mean, na.rm = TRUE)

# Merge total and new_data based on the label column
merged_df1 <- merge(new_data, average_GDP_by_country, by.x = "label", by.y = "regions", all.x = TRUE)

names(merged_df1)[1] <- "regions"

final_df = merge(merged_df, merged_df1, by = 'regions', all= TRUE)

names(final_df)[3] <- "Average_GDP"


ggplot(final_df, aes(x=Average_GDP, y=Average_Price, color=regions)) + geom_point(alpha = 12) + scale_size(range = c(5, 10))
```
The figure above displays the simplified scatter plot for the European countries for GDP and Average price levels. This time each dot represents the respective GDP and average price level for each country in a specific region averaged over all 10 the years. The highest GDP and average price level corresponds to Western Europe and the lowest GDP and average price level.


```{r, eval=FALSE, include=FALSE, echo=FALSE}
##merge population
new_data1 <- data.frame(label = unique(total$regions))

# Calculate average population levels for each country
average_pop_by_country <- aggregate(Population ~ regions, data = total, FUN = mean, na.rm = TRUE)

# Merge total and new_data based on the label column
merged_df2 <- merge(final_df, average_pop_by_country, by="regions", all = TRUE)

names(merged_df2)[1] <- "regions"

final_df1 = merge(merged_df2, final_df, by = 'regions', all= TRUE)

names(final_df1)[3] <- "Average_GDP"

```

```{r}
ggplot(final_df1, aes(x=Average_GDP, y=Average_Price.x, size = Population, color=regions)) + geom_point(alpha = 0.5) + scale_size(range = c(5, 10))
```

